---
title: "Heat Island Analysis of Rochester, New York"
author: Damien Hotte
subtitle: Quantifying the Surface Urban Heat Island (SUHI) effect in Rochester and exploring its relationship with socioeconomic features.
date: today
date-format: long 
---

# Introduction

\[\~ 200 words\]

The Urban Heat Island (UHI) effect is a global phenomenon in which urban areas experience significantly higher Land Surface Temperatures (LST) than their surrounding rural counterparts. This temperature disparity is caused primarily by the dense concentration of buildings, roads, and other impervious surfaces that absorb and retain heat, along with a decrease of vegetation and green spaces that would otherwise provide natural cooling, reduced by urban expansion.

In cities like Rochester, New York, heat patterns do not impact all communities equally. Variations in land use, socio-economic conditions, and access to vegetation can create localized “hot spots” that exacerbate health and environmental inequities. Prolonged exposure to elevated temperatures can increase energy demands, worsen air quality, and put vulnerable populations at risk. This project seeks to examine the relationship between socio-economic and land-use patterns and disparities in heat exposure across Rochester. 

Specifically, it aims to explore how differences in vegetative cover mediate these temperature variations and whether neighborhoods with lower income or less green space experience greater heat stress.By analyzing spatial data on LST, land use, and vegetation distribution, this study will shed light on the socio-environmental dynamics of urban heat and contribute to more equitable strategies for climate resilience.

Note: Changes from brainstorm- narrowing to just focus on Rochester instead of including buffalo, in order to better focus on the comparison of land use and poverty’s impact on green space in Rochester. 


# Materials and methods

\[\~ 200 words\]

Narrative: Clear narrative description of the data sources and methods. Includes data from at least two sources that were integrated / merged in R.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. If you want to use your own data, you must make it available on a website (e.g. Figshare) so that others are able to re-run your code.



```{r, message=F, warning=F}
library(tidyverse)
library(leaflet)
library(kableExtra)
library(htmlwidgets)
library(widgetframe)
library(tidycensus)
library(sf)
library(ggplot2)
library(tigris)
knitr::opts_chunk$set(widgetframe_widgets_dir = 'widgets' ) 
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Download and clean all required data

```{r, message=F, warning=F}
#aquire census tracts 

census_api_key("a7f41abd86b507f92feb140131bdcbd24b0d9170", install = TRUE, overwrite = TRUE) 
readRenviron("~/.Renviron")  

# Set the variable FOR median household income
variables <- c(median_income = "B19013_001")

# Get data for Monroe County
tracts_monroe <- get_acs(
  geography = "tract",
  variables = variables,
  state = "NY",
  county = "Monroe",
  geometry = TRUE
)


rochester_city <- places(state = "NY", cb = TRUE) %>% #tigris places command gets city boundary
  filter(NAME == "Rochester")
tracts_rochester <- tracts_monroe[st_within(tracts_monroe, rochester_city, sparse = FALSE), ] #removes tracts not within 

```
```{r, results='asis'}

map <- ggplot(tracts_rochester) + 
  geom_sf(aes(fill = estimate), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Median Income ($)"
  ) +
  labs(
    title = "Median Household Income by Census Tract",
    subtitle = "Rochester, NY (ACS Data)",
    caption = "Source: U.S. Census Bureau via tidycensus"
  ) +
  theme_minimal()
map
```




# Results

\[\~200 words\]

# Conclusions

\[\~200 words\]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner
